---
title: "EDA"
author: "Andrew Harvey"
format: html
editor: visual
---

# Exploritory Data Analysis File

## Introduction Section

|   This project will use the "diabetes_binary_health_indicators_BRFSS2015" data set. According [Kaggle.com](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data) this data set contains 253,680 observation where the response variable is whether the observed individual does or does not not have diabetes (0 or 1 respectively) or has prediabetes/diabetes (1). The original data set contains 22 variables most of which are categorical. We have limited the variables to those below:

1.  Diabetes_binary (response variable) - This variable illustrates if the subject has or does not have diabetes. 
2.  HighBP (Categorical) - This is a binary variable describing whether or not the subject has high blood pressure.
3.  HighChol (Categorical) - This is a binary variable describing whether or not the subject has high cholesterol.
4.  Smoker (Categorical) - This is a binary variable describing whether or not the subject has smoked at least 100 cigarettes in their lifetime. 100 cigarettes is equal to 5 packs.
5.  HeartDiseaseorAttack (Categorical) - This is a binary variable describing whether or not the subject has coronary heart disease or myocardial infarction (a heart attack).  
6.  PhysActivity (Categorical) - This is a binary variable describing whether or not the subject has conducted physical activity (outside of their job) in the past 30 days.
7.  HvyAlcoholConsump (Categorical) - This is a binary variable describing whether or not the subject, if male, consumes 14 or more alcoholic drinks per week, or if female, consumes 7 or more alcoholic drinks per week. 
8.  MentHlth - This variable is a scale from 1 to 30 of how many days the subject has had POOR mental health.
9.  Age (Categorical) - This variable breaks age into 13 levels which are 18-24, 25-29, 30-34, 35-39, 40-44, 45-49, 50-54, 55-59, 60-64, 65-69, 70-74, 75-80, and 80 or older.

### Exploritory Data Analysis

| The purpose of my exploratory data analysis (EDA) was to identify variables that may have some significant relationship to the diabetes_binary variable. That is to say, does a particular variable either have a relationship to the response variable or not. I explored this by making 4 plots that helped illustrate the possible relationship. I also investigated variables that I did not ultimately use and added variables that I did not investigate initially. The ultimate goal of my model will be to determine what combination of the selected variables are accurate indicators of diabetes. 

## Data
|    First we will load in the packages and the raw data. We then identified categorical variables and their levels by finding unique observations for each variable. Checking for missing values will come next. Finally, variables were converted to factors in order to better display information in the plots. Each plot has a title and a few other visualization adjustments to help the viewer understand. 

```{r Libraries and Raw data, warning=FALSE, message=FALSE,}
library(tidyverse)
```

```{r Data exploration}
rawData <- read.csv("./diabetes_binary_health_indicators_BRFSS2015.csv")

# Quick determination of categorical variables
for (i in 1:length(rawData)) {
  print(unique(rawData[i]))
}
```

```{r Convert variables to factors}
#create a new data frame for cleaning and converts all binary variables into factors
Cleaning1<- rawData|>
  drop_na()|>
  mutate(Diabetes_binary = as.factor(Diabetes_binary),
         HighBP = as.factor(HighBP),
         HighChol = as.factor(HighChol),
         CholCheck = as.factor(CholCheck),
         Smoker = as.factor(Smoker),
         Stroke = as.factor(Stroke),
         HeartDiseaseorAttack = as.factor(HeartDiseaseorAttack),
         PhysActivity = as.factor(PhysActivity),
         Fruits = as.factor(Fruits),
         Veggies = as.factor(Veggies),
         HvyAlcoholConsump = as.factor(HvyAlcoholConsump),
         AnyHealthcare = as.factor(AnyHealthcare),
         NoDocbcCost = as.factor(NoDocbcCost),
         DiffWalk = as.factor(DiffWalk),
         Sex = as.factor(Sex)
         )
```


## Summarizations

| In this section I created 4 summary tables and 4 plots to help illustrate the relation between diabetes_binary and different variables. My initial selection of variables were body mass index (BMI), Mental Health (MentHlth), Heavy Alcohol Consumption (HvyAlcoholConsump), Age, Education, and Income. Ultimately my model used all of these variables except Education and Income.


```{r Summary Tables}
# frequency Body Mass Index (actual) by diabetes_binary
BMIsummary<- Cleaning1|>
  count(Diabetes_binary, BMI, name = "Count")

print(BMIsummary)

ggplot(BMIsummary,aes(x = BMI, y = Count)) +
  labs(title = "BMI Summary by diabetes",
       x = "BMI Value",
       y = "Frequecy")+
  theme(plot.title = element_text(hjust = .5))+
  geom_bar(stat = "identity",position = "dodge", aes(fill = Diabetes_binary))+
  facet_wrap(~Diabetes_binary) 

HealthSummary<-Cleaning1|>
  count(Diabetes_binary,MentHlth)

print(HealthSummary)

ggplot(HealthSummary, aes(MentHlth,n,colour = Diabetes_binary)) +
  labs(title = "Mental Health Summary by diabetes",
       x = "Days of Poor Mental Health",
       y = "Frequecy")+
  theme(plot.title = element_text(hjust = .5))+
  geom_point(stat = "identity")+
  geom_smooth(aes(MentHlth,n),stat = "identity")

# Heavy Alcohol Consumpsion 
# Heavy drinkers (adult men having more than 14 drinks per week and adult women having more than 7 drinks per week)

AlcoholSummary<- Cleaning1|>
  group_by(Diabetes_binary,HvyAlcoholConsump)|>
  tally()

AlcoholSummary

ggplot(Cleaning1,aes(PhysHlth,fill = Diabetes_binary)) +
  geom_bar(position = "dodge") +
  labs(title = "Physical Health Summary by diabetes",
       x = "Days where subject exercised outside of work",
       y = "Frequecy")+
  theme(plot.title = element_text(hjust = .5))


GenBox <- Cleaning1|>
  group_by(Diabetes_binary)|>
  pivot_longer(col = c(Age,Education,Income),names_to = "Category", values_to = "level")|>
  count(Diabetes_binary,Category,level,name = "val")

GenBox

ggplot(GenBox, aes(Diabetes_binary,level, fill = Diabetes_binary)) +
  geom_boxplot()+
  facet_wrap(~Category)+
  scale_y_continuous(breaks = c(1:13))+
  geom_text(aes(label=val))+
  labs(title = "Summary of Age, Education, and Income by diabetes",
       y = "Levels (Values on Plot = Frequency)")+
  theme(plot.title = element_text(hjust = .5))
  

Means<- rawData|>
  select(Diabetes_binary,HighBP,HighChol,HvyAlcoholConsump,Smoker,PhysActivity,Age,HeartDiseaseorAttack,MentHlth,BMI)|>
  summarise(Diabetes_binary.mean = round(mean(Diabetes_binary),digits = 2),
            HighBP.mean = round(mean(HighBP),digits = 2),
            HighChol.mean = round(mean(HighChol),digits = 2),
            HvyAlcoholConsump.mean = round(mean(HvyAlcoholConsump),digits = 2),
            Smoker.mean = round(mean(Smoker),digits = 2),
            PhysActivity.mean = round(mean(PhysActivity),digits = 2),
            AgeGroup.mean = round(mean(Age),digits = 2),
            HeartDiseaseorAttack.mean = round(mean(HeartDiseaseorAttack),digits = 2),
            MentHlth.mean = round(mean(MentHlth),digits = 2),
            BMI.mean = round(mean(BMI),digits = 2))|>
  pivot_longer(cols = 1:10,names_to = "Variable")

Means

ggplot(Means,aes(x=Variable,y=value))+
  geom_bar(stat = "identity", position = "dodge")+ 
  labs(title = "Average by Final Selected Variables",
       x = "Variable",
       y = "Mean")+
  theme(plot.title = element_text(hjust = .5))+
  geom_text(aes(label = value),
            vjust = -.5)+
  theme(axis.text.x = element_text(angle = 15,
                                   vjust = 1.25,
                                   hjust = 1,
                                   size = 8))


```

\`\`\`

[Click here for the Modeling Page](Modeling.html)
